---
title: "Tidy Tuesday"
author: "Daria Patterson"
date: "Nov 4, 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r data-packages}
## Load Data & Packages

library(tidyverse)
library(readr)
library(maps)
library(ggmap)

commute_mode <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")
```


```{r CA-bike}
## What are the top California cities in which people bike to work?

CA_bikecity <- commute_mode %>%
  mutate(city = as.factor(city),
         city = fct_relevel(city, 
          "Menlo Park city", "Berkeley city", "Palo Alto city", 
          "Santa Cruz city", "Davis city")) %>%
  select(city, state_abb, mode, percent) %>%
  filter(state_abb == "CA", mode == "Bike") %>%
  arrange(desc(percent)) %>%
  head(5)
```

```{r CA-bike-plot}
ggplot(data = CA_bikecity, mapping = aes(x = city, y = percent, 
  fill = factor(ifelse(city == "Davis city","Highlighted","Normal")))) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(name = "city", values=c("#DA3934","#3475DA")) +
  coord_flip() +
  labs(x = "City", y = "Percent", title = "Davis: bike-to-work capital of California") +
  theme_minimal()
```

It's interesting that the top five cities in California for which there are the highest percent of people who bike to work are all in Northern California. Moreover, it is facinating that out of these top five cities, four of them (Davis, Berkeley, Santa Cruz) are home to a UC school or distinct private school (Stanford in Palo Alto). Menlo Park, home to Facebook and in the heart of Silicon Valley, is distinct from these other four cities in that it is not an academic hub. 


```{r map-usa}
# create base map of the states in USA

state <- map_data("state")
usa_base <- ggplot(data = state) + 
  geom_polygon(aes(x = long, y = lat, group = group, 
                   fill = region), color = "white") + 
  coord_fixed(1.3) 
```


```{r bike-map-data}
# select the relevant variables, filter for bikers and exclude missing data and states outside of contingent 48.
# caluclate the average percentage of commuters.
# add missing "state" and assign each row a raning, labeled "group"

mean_bike_state <- commute_mode %>%
  select(state, state_abb, percent, mode) %>%
  filter(mode == "Bike", !is.na(state_abb), 
         !(state_abb %in% c("AK", "HI"))) %>%
  group_by(state) %>%
  summarise(mean_percent = mean(percent)) %>%
  add_row(state = "District of columbia", mean_percent = 0) %>%
  arrange(state) %>%
  tibble::rowid_to_column("group")
````


```{r state-clean}
# recode all the group numbers for each state found in the "state" data

state_clean <- state %>%
  mutate(group = case_when(
    region == "massachusetts" ~ 20,
    region == "michigan" ~ 21,
    region == "minnesota" ~ 22,
    region == "mississippi" ~ 23,
    region == "missouri" ~ 24,
    region == "montana" ~ 25,
    region == "nebraska" ~ 26,
    region == "nevada" ~ 27,
    region == "new hampshire" ~ 28,
    region == "new jersey" ~ 29,
    region == "new mexico" ~ 30,
    region == "new york" ~ 31,
    region == "north carolina" ~ 32,
    region == "north dakota" ~ 33,
    region == "ohio" ~ 34,
    region == "oklahoma" ~ 35,
    region == "oregon" ~ 36,
    region == "pennsylvania" ~ 37,
    region == "rhode island" ~ 38,
    region == "south carolina" ~ 39,
    region == "south dakota" ~ 40,
    region == "tennessee" ~ 41,
    region == "texas" ~ 42,
    region == "utah" ~ 43,
    region == "vermont" ~ 44,
    region == "virginia" ~ 45,
    region == "washington" ~ 46,
    region == "west virginia" ~ 47,
    region == "wisconsin" ~ 48,
    region == "wyoming" ~ 49,
    
    TRUE ~ group
  ))

```


```{r mean-map-data}
# join the state df and the df with the calculated mean percentages

mean_map_data <- mean_bike_state %>%
  inner_join(state_clean, by = "group")
mean_map_data
```


```{r no-axes}
#create a theme with no axes, borders, basically a blank slate

no_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )
```


```{r mean-map}
# build map of the average percentage of commuters in each state who bike to work

mean_bike_usa <- ggplot(data = mean_map_data) +
  geom_polygon(aes(x = long, y = lat, fill = mean_percent, group = group), color = "white") +
  labs(title = "Vermont: bike-to-work capital of the US",
       fill = "Average percent commuters \n (per state)") +
  scale_fill_gradientn(colours = rev(rainbow(7)),
                         breaks = c(0, 1, 2, 3, 4, 5)) +
  coord_fixed(1.3) +
  theme_bw() +
  theme(legend.position = "bottom", plot.title = element_text(hjust=0.5)) +
    no_axes
mean_bike_usa
```

It is very surprising that Vermont has the highest average percentage of commuters who bike to work!



